Universal algebra and the \texorpdfstring{$ \lambda $}{lambda}-calculus

### 2024-05-28 report/src/univalent-foundations.tex
Change to the TU Delft PL template

So if we have types and elements, are types also elements of some type? This is a tricky question, because if we say that there exists a type $ \TYPE $ which contains all types (and therefore, itself), we introduce inconsistency in our type system (see \cite{girard}). This is just like having a `set of all sets', which results in problematic questions like ``does its subset, containing only the sets which do not contain itself, contain itself?'' This is usually solved by either stating that types do not themselves have a type, or by assuming the existence of \textit{type universes} $ U_1, U_2, \dots $, such that every type is an element of some $ U_i $. Note that universes are allowed to be inclusive or not: we may have $ U_n \subseteq U_{n + 1} $. In this thesis, we will not explicitly mention particular universes. We will just write $ \TYPE $ to denote some type universe (or its category of types and functions). This is called \textit{typical ambiguity}, meaning that every theorem holds for all (suitable) assignments of universes to these instances of $ \TYPE $.

@@

Now, the \iindex{univalence principle} takes this a step further. It states that
\begin{quote}
  \textit{Isomorphic objects are equal.}
\end{quote}
Univalent foundations seeks to be a foundation for mathematics that is in line with this principle. This is often done within the framework of `Martin-Löf dependent type theory', a type theory developed by Per Martin-Löf \cite{martin-lof-type-theory}. It is a family of dependent type systems with dependent products, dependent sums, an empty type, a unit type and union types. It is a constructive type theory, so it does not automatically assume the axiom of excluded middle or the propositional axiom of choice. It is however compatible with these axioms, so one can still assume these alongside its usual axioms.

It is important to note that Martin-Löf type theory has \textit{identity types}: given a type $ T $ and elements $ x, y: T $, we have a type $ \mathrm{Id}_T(x, y) $ (note that this is a dependent type), which we will usually denote with $ x = y $. An element $ p: x = y $ is a proof that $ x $ is `equal' to $ y $. This type comes with an interesting induction principle called \iindex{path induction}: we can show any statement about paths $ p: x = y $ for generic $ x $ and $ y $, if we can show it about the `trivial' path $ \mathrm{refl}: x = x $ for generic $ x $. For example, symmetry of the equality boils down to a function $ \prod_{x y : T} x = y \to y = x $. We construct this using path induction with the function that sends $ \mathrm{refl}: x = x $ to itself. For more information, see \cite{hottbook}, Section 1.12.1.

In set-theoretic mathematics, there is the concept of a `bijection' $ S \simeq T $ of sets (or an isomorphism in the category $ \SET $), which is often treated as an equivalence. It consists of functions $ f: S \to T $ and $ g: T \to S $ with $ f \cdot g = \id S $ and $ g \cdot f = \id T $. In type theory, we have a similar concept, which is called `equivalence' (of types) $ S \simeq T $. Since bijections are not well-behaved for types that are not sets (see \ref{sec:homotopy-props-sets}), because in those cases, a function $ f: S \to T $ can have multiple distinct inverses. Therefore, we define $ S \simeq T := \sum_{f: S \to T} \mathrm{isequiv}(f) $, for some predicate $ \mathrm{isequiv}: (S \to T) \to \TYPE $ (see \cite{hottbook}, Equation 2.4.10). However, intuitively we can still think of these as bijections.

@@

\begin{lemma}
  For every category $ C $, there exists a univalent category $ \hat C $ with a weak equivalence
  \[ \iota: C \hookrightarrow \hat C. \]
\end{lemma}
\begin{proof}
  See \cite{univalent-categories}, Theorem 8.5.
\end{proof}
We call this univalent category the \iindex{Rezk completion}.
\TODO: Uniqueness?

@@

Also, Theorem 4.5 in \cite{univalent-categories} shows that if a category $ B $ is univalent (in the paper, categories are called `precategories` and univalent categories are just called `categories'), then the functor category $ A \rightarrow B $ is also univalent. In particular, the category of (pre)sheaves $ A \to \SET $ is univalent.

@@

The last structure in this section for which we want to consider the univalence axiom, is the 2-category $ \Cat $ of categories. In general, we cannot show that this satisfies the univalence principle. However, we will restrict our attention to the sub-2-category of univalent categories, which are the categories that we want to study. Then Theorem 6.8 in \cite{univalent-categories} shows that for univalent categories $ C $ and $ D $, there is an equivalence between $ C = D $ and $ C \simeq D $, where $ C \simeq D $ denotes the type of (adjoint) equivalences of categories (see Definition \ref{def:equivalence-of-categories}).

Lastly, a result about univalent categories that we will use a couple of times in this thesis:
\begin{lemma}\label{lem:univalen-category-equivalence}
  For a functor between univalent categories $ F: A \to B $, the types `$ F $ is an adjoint equivalence' and `$ F $ is a weak equivalence' are equivalent propositions (see \ref{sec:homotopy-props-sets}).
\end{lemma}
\begin{proof}
  See \cite{univalent-categories}, Lemma 6.8.
\end{proof}

@@

\begin{definition}
  For a type $ T $, a type $ \Vert T \Vert $ exists (\cite{hottbook}, Section 3.7) with the properties that for all $ t: T $, we have an element $ \vert t \vert : \Vert T \Vert $ and that $ \Vert T \Vert $ is a mere proposition. It has a recursion principle stating that for a mere proposition $ B $, a function $ f: A \to B $ induces a function $ \vert f \vert: \Vert T \Vert \to B $ that commutes with $ \vert \cdot \vert $. This object is called the \index{truncation!propositional}\textit{propositional truncation}.
\end{definition}
The propositional truncation forgets the details about a type, and only keeps the information about whether it is inhabited or not. The recursion principle means that if we are trying to prove a mere proposition based on some element $ \vert t \vert: \Vert T \Vert $, we can pretend that we do have a concrete element $ t: T $.

@@

\begin{figure}
  \begin{subfigure}{.3\textwidth}
    \includesvg[pretex=\tiny, width=\columnwidth]{images/homotopy-plane}
    \caption{}
    \label{fig:homotopy-plane}
  \end{subfigure}
  \begin{subfigure}{.3\textwidth}
    \includesvg[pretex=\tiny, width=\columnwidth]{images/homotopy-cylinder}
    \caption{}
    \label{fig:homotopy-cylinder}
  \end{subfigure}
  \begin{subfigure}{.3\textwidth}
    \includesvg[pretex=\tiny, width=\columnwidth]{images/homotopy-sphere}
    \caption{}
    \label{fig:homotopy-sphere}
  \end{subfigure}
  \caption{Different possible homotopy structures}
\end{figure}

@@

  For something to look like a circle or tube, we need this group $ x = x $ to be isomorphic to $ \mathbb Z $. For an example, see \cite{circle-fundamental-group}.
\end{remark}

@@

\begin{figure}
  \begin{subfigure}{.45\textwidth}
    \includesvg[pretex=\tiny, width=.7\columnwidth]{images/fibration}
    \caption{}
    \label{fig:fibration}
  \end{subfigure}
  \begin{subfigure}{.45\textwidth}
    \includesvg[pretex=\tiny, width=.7\columnwidth]{images/transport}
    \caption{}
    \label{fig:transport}
  \end{subfigure}
  \caption{A fibration with a path in the base space, giving rise to transport in the fibration}
\end{figure}

>>>>

So if we have types and elements, are types also elements of some type? This is a tricky question, because if we say that there exists a type $ \TYPE $ which contains all types (and therefore, itself), we introduce inconsistency in our type system (see \textcite{girard}). This is just like having a `set of all sets', which results in problematic questions like ``does its subset, containing only the sets which do not contain itself, contain itself?'' This is usually solved by either stating that types do not themselves have a type, or by assuming the existence of \textit{type universes} $ U_1, U_2, \dots $, such that every type is an element of some $ U_i $. Note that universes are allowed to be inclusive or not: we may have $ U_n \subseteq U_{n + 1} $. In this thesis, we will not explicitly mention particular universes. We will just write $ \TYPE $ to denote some type universe (or its category of types and functions). This is called \textit{typical ambiguity}, meaning that every theorem holds for all (suitable) assignments of universes to these instances of $ \TYPE $.

@@

Now, the \iindex{univalence principle} takes this a step further. It states that
\begin{quote}
  \textit{Isomorphic objects are equal.}
\end{quote}
Univalent foundations seeks to be a foundation for mathematics that is in line with this principle. This is often done within the framework of `Martin-Löf dependent type theory', a type theory developed by Per Martin-Löf \textcite{martin-lof-type-theory}. It is a family of dependent type systems with dependent products, dependent sums, an empty type, a unit type and union types. It is a constructive type theory, so it does not automatically assume the axiom of excluded middle or the propositional axiom of choice. It is however compatible with these axioms, so one can still assume these alongside its usual axioms.

It is important to note that Martin-Löf type theory has \textit{identity types}: given a type $ T $ and elements $ x, y: T $, we have a type $ \mathrm{Id}_T(x, y) $ (note that this is a dependent type), which we will usually denote with $ x = y $. An element $ p: x = y $ is a proof that $ x $ is `equal' to $ y $. This type comes with an interesting induction principle called \iindex{path induction}: we can show any statement about paths $ p: x = y $ for generic $ x $ and $ y $, if we can show it about the `trivial' path $ \mathrm{refl}: x = x $ for generic $ x $. For example, symmetry of the equality boils down to a function $ \prod_{x y : T} x = y \to y = x $. We construct this using path induction with the function that sends $ \mathrm{refl}: x = x $ to itself. For more information, see \textcite{hottbook}, Section 1.12.1.

In set-theoretic mathematics, there is the concept of a `bijection' $ S \simeq T $ of sets (or an isomorphism in the category $ \SET $), which is often treated as an equivalence. It consists of functions $ f: S \to T $ and $ g: T \to S $ with $ f \cdot g = \id S $ and $ g \cdot f = \id T $. In type theory, we have a similar concept, which is called `equivalence' (of types) $ S \simeq T $. Since bijections are not well-behaved for types that are not sets (see \ref{sec:homotopy-props-sets}), because in those cases, a function $ f: S \to T $ can have multiple distinct inverses. Therefore, we define $ S \simeq T := \sum_{f: S \to T} \mathrm{isequiv}(f) $, for some predicate $ \mathrm{isequiv}: (S \to T) \to \TYPE $ (see \textcite{hottbook}, Equation 2.4.10). However, intuitively we can still think of these as bijections.

@@

\begin{lemma}
  For every category $ C $, there exists a univalent category $ \hat C $ with a weak equivalence
  \[ \iota: C \hookrightarrow \hat C. \]
\end{lemma}
\begin{proof}
  See \textcite{univalent-categories}, Theorem 8.5.
\end{proof}
We call this univalent category the \iindex{Rezk completion}.
\TODO: Uniqueness?

@@

Also, Theorem 4.5 in \textcite{univalent-categories} shows that if a category $ B $ is univalent (in the paper, categories are called `precategories` and univalent categories are just called `categories'), then the functor category $ A \rightarrow B $ is also univalent. In particular, the category of (pre)sheaves $ A \to \SET $ is univalent.

@@

The last structure in this section for which we want to consider the univalence axiom, is the 2-category $ \Cat $ of categories. In general, we cannot show that this satisfies the univalence principle. However, we will restrict our attention to the sub-2-category of univalent categories, which are the categories that we want to study. Then Theorem 6.8 in \textcite{univalent-categories} shows that for univalent categories $ C $ and $ D $, there is an equivalence between $ C = D $ and $ C \simeq D $, where $ C \simeq D $ denotes the type of (adjoint) equivalences of categories (see Definition \ref{def:equivalence-of-categories}).

Lastly, a result about univalent categories that we will use a couple of times in this thesis:
\begin{lemma}\label{lem:univalen-category-equivalence}
  For a functor between univalent categories $ F: A \to B $, the types `$ F $ is an adjoint equivalence' and `$ F $ is a weak equivalence' are equivalent propositions (see \ref{sec:homotopy-props-sets}).
\end{lemma}
\begin{proof}
  See \textcite{univalent-categories}, Lemma 6.8.
\end{proof}

@@

\begin{definition}
  For a type $ T $, a type $ \Vert T \Vert $ exists (\textcite{hottbook}, Section 3.7) with the properties that for all $ t: T $, we have an element $ \vert t \vert : \Vert T \Vert $ and that $ \Vert T \Vert $ is a mere proposition. It has a recursion principle stating that for a mere proposition $ B $, a function $ f: A \to B $ induces a function $ \vert f \vert: \Vert T \Vert \to B $ that commutes with $ \vert \cdot \vert $. This object is called the \index{truncation!propositional}\textit{propositional truncation}.
\end{definition}
The propositional truncation forgets the details about a type, and only keeps the information about whether it is inhabited or not. The recursion principle means that if we are trying to prove a mere proposition based on some element $ \vert t \vert: \Vert T \Vert $, we can pretend that we do have a concrete element $ t: T $.

@@

\begin{figure}
  \begin{subfigure}{.3\textwidth}
    \includesvg[pretex=\tiny, width=\columnwidth]{img/homotopy-plane}
    \caption{}
    \label{fig:homotopy-plane}
  \end{subfigure}
  \begin{subfigure}{.3\textwidth}
    \includesvg[pretex=\tiny, width=\columnwidth]{img/homotopy-cylinder}
    \caption{}
    \label{fig:homotopy-cylinder}
  \end{subfigure}
  \begin{subfigure}{.3\textwidth}
    \includesvg[pretex=\tiny, width=\columnwidth]{img/homotopy-sphere}
    \caption{}
    \label{fig:homotopy-sphere}
  \end{subfigure}
  \caption{Different possible homotopy structures}
\end{figure}

@@

  For something to look like a circle or tube, we need this group $ x = x $ to be isomorphic to $ \mathbb Z $. For an example, see \textcite{circle-fundamental-group}.
\end{remark}

@@

\begin{figure}
  \begin{subfigure}{.45\textwidth}
    \includesvg[pretex=\tiny, width=.7\columnwidth]{img/fibration}
    \caption{}
    \label{fig:fibration}
  \end{subfigure}
  \begin{subfigure}{.45\textwidth}
    \includesvg[pretex=\tiny, width=.7\columnwidth]{img/transport}
    \caption{}
    \label{fig:transport}
  \end{subfigure}
  \caption{A fibration with a path in the base space, giving rise to transport in the fibration}
\end{figure}
